\documentclass{article}
\usepackage[margin=1in]{geometry}

\usepackage[T1]{fontenc}

\usepackage{color}
\definecolor{lightgray}{rgb}{0.95, 0.95, 0.95}
\definecolor{darkgray}{rgb}{0.4, 0.4, 0.4}
\definecolor{purple}{rgb}{0.65, 0.12, 0.82}
\definecolor{editorGray}{rgb}{0.95, 0.95, 0.95}
\definecolor{editorOcher}{rgb}{1, 0.5, 0} % #FF7F00 -> rgb(239, 169, 0)
\definecolor{editorGreen}{rgb}{0, 0.5, 0} % #007C00 -> rgb(0, 124, 0)

\usepackage{titling}
\title{CS 5300 Project \#1}
\author{Jared Rainwater \& Samuel K. Grush}

\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\lhead{\thetitle}
\rhead{\theauthor}
\cfoot{Page \thepage}

\usepackage{upquote}
\usepackage{listings}
\input{lstdefines}

\lstset{%
    % Basic design
    backgroundcolor=\color{editorGray},
    basicstyle={\small\ttfamily},
    frame=l,
    % Line numbers
    xleftmargin={0.75cm},
    numbers=left,
    stepnumber=1,
    firstnumber=1,
    numberfirstline=true,
    % Code design
    keywordstyle=\color{blue}\bfseries,
    commentstyle=\color{darkgray}\ttfamily,
    ndkeywordstyle=\color{editorGreen}\bfseries,
    stringstyle=\color{editorOcher},
    % Code
    language=TypeScript,
    alsodigit={.:;},
    tabsize=2,
    showtabs=false,
    showspaces=false,
    showstringspaces=false,
    extendedchars=true,
    breaklines=true,
}

\newcommand{\filelisting}[2][TypeScript]{%
  \subsubsection{\texttt{\protect#2}}
  \lstinputlisting[language=#1, label={lst:#2}]{../#2}
  \pagebreak[3]\-\\
}


\begin{document}

\maketitle

\tableofcontents

\section{The Compiler}

In order to parse SQL commands, we are using a parsing library called \textbf{PEG.js},
which allows us to express a/n SQL syntax as a \emph{Parsing Expression Grammar} (PEG),
and build that grammar into a JavaScript parser.
The grammar was initially structured after Phoenix's SQL grammar,
but generally follows PostgreSQL's syntax and the corresponding ANSI SQL standard.

\subsection{Grammar Rules}

\emph{The grammar is defined in} \verb|src/parser/peg/sql.pegjs|. \\

Parsing starts out with the \verb|Statements| rule, which is a semicolon
delimited list of SQL \verb|Statement|s.
A \verb|Statement| can be either a \verb|Select| or \verb|SelectPair|.
\verb|Select| is broken up into 6 clauses: \verb|TargetClause|, \verb|FromClause|,
\verb|WhereClause|, \verb|GroupByClause|, \verb|HavingClause| and \verb|OrderByClause|.
These correspond to all the possibilities of a valid SQL Select statement.
A \verb|SelectPair| is two seperate \verb|Select| clauses paired together with a
``UNION'', ``INTERSECT'', or ``EXCEPT'' set operation.
You can also apply the ``ALL'' or ``DISTINCT'' modifier to the pair.

The \verb|TargetClause| can have the optional ``DISTINCT'' or ``ALL'' modifier
followed by ``*'' (to allow everything) or a \verb|TargetList|, a
comma-delimited list of \verb|TargetItem|s.
A \verb|TargetItem| is a column-like specifier; it can be a relation name with ``.*''
or an \verb|Operand| with optional alias.

\verb|FromClause| aliases \verb|RelationList|, a list of comma-delimited
relation-like fields, each of which may be a table name (with optional alias)
or a \verb|Join|. A \verb|JOIN| is a pair of relation-like fields joined by
a join-type (``CROSS'', ``INNER'', ``LEFT'', etc) followed by an optional
join-condition (``ON \verb|Condition|'' or ``USING (\verb|TargetList|)'').

\verb|WhereClause| and \verb|HavingClause| are \verb|Condition|s.
The types of \verb|Condition|s are: ``OR'' and ``AND'' (which join two
\verb|Condition|s); comparison, ``LIKE'', and ``BETWEEN'' (which join two
\verb|Operands|); and ``IN'' and ``EXISTS'' (which take \verb|Select|-like
arguments).

\verb|GroupByClause| is simply a \verb|TargetList| like the target clause.
\verb|OrderByClause| is a comma-delimited list of \verb|Operand|s, each optionally
with an ordering-condition (``ASC'', ``DESC'' ``USING \ldots'').

An \verb|Operand| is a \verb|Term| optionally joined to other \verb|Operand|s
by value operations (e.g. arithmetic or concatenation).
A \verb|Term| is a \verb|Literal|, aggregate function, or column reference.
\verb|Literal|s include numeric literals, booleans literals, and string literals
(single-quoted).

A \verb|Name|, which might refer to an operand or relation, is denoted by a
bare-identifier (\verb|/[a-z_][a-z0-9_]*/| and not a \verb|ReservedWord|) or any
string quoted with double-quotes (\verb|"..."|) or backticks (\verb|`...`|).

Both comment forms are supported: starting with \verb|--| and consuming the
rest of the line, and C-style starting with \verb|/*| and ending at \verb|*/|.
Both are permitted anywhere whitespace is.

The \verb|ReservedWord| rule contains 340 keywords that the ISO/ANSI SQL:2008
standard states are \textbf{never allowed as identifiers}. This set is almost
certainly overkill, as most SQL implementations only reserve a \emph{small}
fraction of it. It is also excessively large, making up over $^1/_3$ of the
grammar's sourcecode and $\textbf{90\%}$ of the uncompressed compiled grammar.

\subsection{Interpretation}

\emph{Classes and data structures discussed in this section defined in}
      \verb|src/parser/types.ts|. \\

While parsing the grammar, the PEG.js parser calls JavaScript classes that
correspond to SQL concepts.
These classes include \verb|SqlSelect|, \verb|SqlJoin|, \verb|SqlConditional|,
\verb|SqlLiteral|, etc.
This generates an object-oriented data structure---resembling a tree---that
represents the ``SQL Structure''.

Once the SQL Structure is generated it can be converted into JavaScript classes
that correspond to Relational Algebra concepts.
These classes include \verb|RelRestriction|, \verb|RelProjection|,
\verb|RelJoin|, \verb|RelConditional|, etc.
This generates a data structure---more closely resembling a tree than before---
that represents the ``Relational Algebra Structure''.

\emph{Top-level functions for parsing/conversion defined in} \verb|src/parser/parsing.ts|,
\emph{with conversion implementation functions defined in} \verb|src/parser/sqlToRel.ts|.

\pagebreak

\subsection{Optimization}

Once a query has been processed, it is ready for execution. But performance
gains are possible by further analyzing the quey and restructuring it.
This is an optimization process that works by looking at the structure and
operations  of the query and modifying it based on specific rules and
heuristics. All of the optimization are based on specific transformations.

\subsubsection{Transformation Rules}

These transformations are based on those described in Elmasri on page 698--699.

\begin{enumerate}
  \item \textbf{cascading of restrictions:} If a restriction is made that has multiple conditions anded together,
  an individual restrict can be made and they can be nested within one another.

  \item \textbf{commute restriction:} If a restrictions operates directly on another restriction nested within it, then the order of
  the restrictions can be changed.

  \item \textbf{cascade selection:} In a sequence of selection operations, all but the last one can be ignored.

  \item \textbf{commute restriction with selection:} If a selection deals only with attributes pertaining to selection,
  the restriction and selection are commutative.

  \item \textbf{commute cross product or join:} The relations associated with either the cross product or join
  can have their order changed.

  \item \textbf{commute selection with join/cross product:} A selection on the result of a join or cross product can move to just
  an individual relation if its conditions only apply to attributes from the relation.

  \item \textbf{Commute restriction with join/cross product:} A restriction that operates on the result of a join or cross product
  can be split into 2 restrictions operating on either argument. Either restrictions' columns will now pertain only
  to the relation it acts on.

  \item \textbf{commute set operations:} union and intersection are commutative.

  \item \textbf{associativity of join, cross product, union and intersection:} the mentioned operations are associative with themselves

  \item \textbf{selection commutes with set operations:} for any set operation, a selection operating on its result is the same as
  it operating on the arguments

  \item \textbf{restriction commutes with union:} a restriction on a union is the same as restricting the arguments of the union

  \item \textbf{convert selection on a cross product to join:} When using a selection on a cross product, you can convert it
  to a join with its condition matching the selection.

  \item \textbf{selection in conjuction with set minus:} If commuting a selection with set minus, the select only moves to the first,
  left hand argument.

  \item \textbf{selection in conjuction with intersect:} if a selection acting on an intersection has conditions pertaining
  only to 1 of the arguments, you can apply the selection to just that argument
\end{enumerate}


\pagebreak

\section{Source Code}

{\large \textbf{All of this code is available at} \verb|https://github.com/SKGrush/sqlparse5300|}

\subsection{\texttt{src/}}

\filelisting[TypeScript]{src/index.ts}
\filelisting[ReactTypeScript]{src/Main.tsx}

\subsection{\texttt{src/components}}

\filelisting[ReactTypeScript]{src/components/QueryInput.tsx}
\filelisting[ReactTypeScript]{src/components/RelationsInput.tsx}
\filelisting[ReactTypeScript]{src/components/TestCase.tsx}
\filelisting[ReactTypeScript]{src/components/Tests.tsx}
\filelisting[ReactTypeScript]{src/components/tree.tsx}

\subsection{\texttt{src/parser}}

\filelisting{src/parser/parsing.ts}
\filelisting{src/parser/relAnalysis.ts}
\filelisting[ReactTypeScript]{src/parser/relationalText.tsx}
\filelisting{src/parser/relDupe.ts}
\filelisting{src/parser/relTransform.ts}
\filelisting{src/parser/sqlToRel.ts}
\filelisting{src/parser/tests.ts}

\subsection{\texttt{src/parser/peg}}

\filelisting[pegjs]{src/parser/peg/sql.pegjs}
\filelisting[pegjs]{src/parser/peg/relations.pegjs}

\subsection{\texttt{src/parser/types}}

\filelisting{src/parser/types/index.ts}
\filelisting{src/parser/types/Catalog.ts}
\filelisting{src/parser/types/Rel.ts}
\filelisting{src/parser/types/Sql.ts}

\subsection{\texttt{src/query\_tree}}

\filelisting[TypeScript]{\detokenize{src/query_tree/node.ts}}
\filelisting[TypeScript]{\detokenize{src/query_tree/operation.tsx}}

\end{document}
